---
title: "How do literacy and age of a marriage affect family size?"
author: "Maryam Ansari & Alexia Xu"
date: January 27, 2025
output: pdf_document
toc: true
bibliography: references.bib
---
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r libraries, include=FALSE}
library(tidyr)
library(kableExtra)
library(tidyverse)
library(ggplot2)
```


```{r dataDownload, echo= FALSE, include=FALSE}
pUrl = 'http://wfs.dhsprogram.com/pt/ptsr01.dat'
pName = file.path(tempdir(), 'portugal.dat')
if(!file.exists(pName)) {
  download.file(pUrl, pName)
}

datNames = rbind(
		age=c(45,2),
		ageMarried=c(149,2), 
		monthsSinceM = c(157,4),
#		failedPregnancies=c(421,2),
#		failedPregStill=c(423,2),
#		failedPregSpAb=c(425,2),
		pregnancies=c(433,2),
		children=c(435,2),
		sons=c(443,2),
#		firstBirthInterval = c(479,2),
		region = c(641,2),
		literacy = c(649,2)
)
		colnames(datNames ) = c('start','len')
		datNames = cbind(startm1=datNames[,1]-1,datNames, sum=apply(datNames, 1,sum))
		cbind(datNames[-1,1] , datNames[seq(1, nrow(datNames)-1),4])
		datNames[-1,1] = datNames[-1,2] - datNames[seq(1, nrow(datNames)-1),4]
		dWidths = as.vector(t(datNames[,c(1,3)]))
		dNames = paste(rep(rownames(datNames), rep(2, nrow(datNames))),
  	rep(c( "junk",""), nrow(datNames)), sep="") 
		
		dNames = dNames[dWidths > 0]
		dWidths = dWidths[dWidths > 0]
		
		formats = list(
			ageMarried = data.frame(code=1:7,  label=c(0,15,18,20,22,25,30)),
			region = data.frame(code=1:5, 
				label=c('lisbon','porto','20k+', '10-20k', 'lt10k')),
			literacy = data.frame(code=1:2, label=c('yes','no')),
			firstBirthInterval = data.frame(
					code = 1:8,
					label = c(
							'lt0','0-7', '8-11','12-23',
							'24-35','36-47','48-59','60-Inf'
							)
					)
		)

	

		formats$ageMarried$label = 
  	paste(formats$ageMarried$label, 'to',
  	c(formats$ageMarried$label[-1], 'Inf'), sep='')
  	formats$ageMarried = rbind(formats$ageMarried, data.frame(code=88, label='never'))

   
  portugal = read.fwf(
    pName,
    dWidths, col.names=dNames,
    header=FALSE)
  
  portugal = portugal[,grep("junk$", names(portugal), invert=TRUE)]

for(D in intersect(names(portugal), names(formats))){
  		portugal[[D]] = factor(portugal[[D]],
  			levels=formats[[D]]$code, 
				labels=formats[[D]]$label)
}
portugal$ageMarried = relevel(portugal$ageMarried, '22to25')
portugal$region = relevel(portugal$region, 'lt10k')

if(FALSE) save(portugal, file='portugal.RData')
```

```{r thedata, echo=FALSE}
head(portugal)
table(portugal$region)
```

\newpage

# Introduction

## Checklist

- Strong and relevant motivation as to why it is important to study this particular research question
- Summaries of the main results from three peer-reviewed articles on same or relevant topic
- Make sure that articles talk about which variables have been shown to be related to the response variables in your data.


## Actual Text




\newpage

# Methods

## Checklist

- Choice of GLM model that will be fit
- Justification for model based on data and research goal
- Process for decision on the appropriateness of the model.

## Planning

- Categorical: age, ageMarried, region
- Binary: Literacy
- Response: children -> count variable

- Poisson should be best since response is a count variable?
- Maybe neg binomial?

```{r EDA, include=FALSE}
# Distribution of family size
portugal %>%
  ggplot(aes(x = children)) + 
  geom_bar(fill = "steelblue") + 
  labs(title = "Distribution of Family Size", x = "Number of Children", y = "Frequency")

# Literacy and family size
ggplot(portugal, aes(x = literacy, y = children)) + 
  geom_boxplot(fill = "lightgreen") + 
  labs(title = "Family Size by Literacy", x = "Literacy", y = "Family Size")

```

Fitting the model:
```{r model, include=FALSE}
# Fit a Poisson model
poisson_model <- glm(children ~ literacy + ageMarried + region, 
                     family = poisson(link = "log"), data = portugal)

# Summary of the model
summary(poisson_model)

```

Checking for overdispersion:
```{r}
# Check overdispersion
residual_deviance <- deviance(poisson_model)
df_residual <- df.residual(poisson_model)
overdispersion_ratio <- residual_deviance / df_residual

cat("Residual Deviance:", residual_deviance, "\n")
cat("Degrees of Freedom:", df_residual, "\n")
cat("Overdispersion Ratio:", overdispersion_ratio, "\n")

```

**ChatGPT interpretation of overdispersion ratio** 
- If the overdispersion ratio is close to 1, the Poisson model is a good fit for the data, as the variance roughly equals the mean (an assumption of the Poisson distribution).
- A ratio of 1.177 is slightly greater than 1, suggesting mild overdispersion (the variance is slightly greater than the mean). 


Residual Analysis:
```{r}
# Residual diagnostics
par(mfrow = c(2, 2))
plot(poisson_model)
```


## Actual Text



\newpage

# Results

## Checklist

- Statistical summaries of the data and variables being considered 
- Discussion of the data distributions and their preliminary implications regarding the research question.
- Presentation of model results (e.g. estimated coefficients, standard errors, CIs) and evidence supporting
inclusion/exclusion of predictors.
- Discussion of the process and results involved in developing a suitable model that addresses the
research goal

## Actual Text



\newpage

# Conclusion

## Checklist

- A formal statistical interpretation of a coefficient of at minimum one of the most important or relevant
predictors.
- A general summary of what your model as a whole says about the research question.
- A comparison to the literature results summarized in the introduction section, commenting on what is
similar and what is different.



## Actual Text





\newpage

# References

